---
title: "Hands-on Exercise 1"
author: "Walaiporn Pawitranon"
---

# Getting Started

## Install and launching R packages

The code chunk below use p_load function of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will bbe launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Plotting a simple chart

## A bar chart

The code chunk below plots a bar chart by using [`geom_bar()`](https://ggplot2.tidyverse.org/reference/geom_bar.html).

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```

## A dot plot

In the code chunk below, [`geom_dotplot()`](https://ggplot2.tidyverse.org/reference/geom_dotplot.html) of ggplot2 is used to plot a dot plot.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

The code chunk below is used to scale the y axis

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL) 
```

## A simple histogram

In the code chunk below, geom_histogram() is used to create a simple histogram by using values in MATHS field of exam_data.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()
```

### Modifying a geometric object by changing geom()

In the code chunk below,

-   *bins* argument is used to change the number of bins to 20,

-   *fill* argument is used to shade the histogram with light blue color, and

-   *color* argument is used to change the outline colour of the bars in black

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")
```

### Modifying a geometric object by changing aes()

The code chunk below changes the interior colour of the histogram (i.e. *fill*) by using sub-group of *aesthetic()*.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

## A kernel density estimate

A smoother version of histogram. The code below plots the distribution of Maths scores in a kernel density estimate plot.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density() 
```

The code chunk below plots two kernel density lines by using *colour* or *fill* arguments of *aes()*

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

## A box plot

The code chunk below plots boxplots by using [`geom_boxplot()`](https://ggplot2.tidyverse.org/reference/geom_boxplot.html).

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()             
```

Notches are used in box plots to help visually assess whether the medians of distributions differ. If the notches do not overlap, this is evidence that the medians are different.

The code chunk below plots the distribution of Maths scores by gender in notched plot instead of boxplot.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)            
```

## A violin plot

[`geom_violin`](https://ggplot2.tidyverse.org/reference/geom_violin.html) is designed for creating violin plot. Violin plots are a way of comparing multiple data distributions. With ordinary density curves, it is difficult to compare more than just a few distributions because the lines visually interfere with each other. With a violin plot, it’s easier to compare several distributions since they’re placed side by side.

The code below plot the distribution of Maths score by gender in violin plot.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()           
```

A scatter plot

The code chunk below plots a scatterplot showing the Maths and English grades of pupils by using geom_point().

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()           
```

A combined plot

The code chunk below plots the data points on the boxplots by using both geom_boxplot() and geom_point().

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)                   
```

# The statistic functions

The [Statistics functions](https://ggplot2.tidyverse.org/reference/index.html#stats) statistically transform data, usually as some form of summary. For example:

-   frequency of values of a variable (bar graph)

    -   a mean

    -   a confidence limit

-   There are two ways to use these functions:

    -   add a `stat_()` function and override the default geom, or

    -   add a `geom_()` function and override the default stat.

## The statistic summary

### The stat_summary() method

The code chunk below adds mean values by using [`stat_summary()`](https://ggplot2.tidyverse.org/reference/stat_summary.html) function and overriding the default geom.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)                                  
```

### The geom() method

The code chunk below adding mean values by using `geom_()` function and overriding the default stat.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)                                          
```

## Adding best fit curve on a scatter plot

The scatterplot below shows the relationship of Maths and English grades of pupils. The interpretability of this graph can be improved by adding a best fit curve.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()           
```

In the code chunk below, [`geom_smooth()`](https://ggplot2.tidyverse.org/reference/geom_smooth.html) is used to plot a best fit curve on the scatterplot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)           
```

The default smoothing method can be overridden as shown below.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)         
```

# Facets

Facetting generates small multiples (sometimes also called trellis plot), each displaying a different subset of the data. They are an alternative to aesthetics for displaying additional discrete variables. ggplot2 supports two types of factes, namely: [`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html) and [`facet_wrap`](https://ggplot2.tidyverse.org/reference/facet_wrap.html).

## Working with facet_wrap()

[`facet_wrap`](https://ggplot2.tidyverse.org/reference/facet_wrap.html) wraps a 1d sequence of panels into 2d. This is generally a better use of screen space than facet_grid because most displays are roughly rectangular.

The code chunk below plots a trellis plot using `facet-wrap()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

## facet_grid() function

[`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html) forms a matrix of panels defined by row and column facetting variables. It is most useful when you have two discrete variables, and all combinations of the variables exist in the data.

The code chunk below plots a trellis plot using `facet_grid()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS) 
```

# Coordinates functions

The *Coordinates* functions map the position of objects onto the plane of the plot.

## Working with Coordinate

By the default, the bar chart of ggplot2 is in vertical form.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() 
```

The code chunk below flips the horizontal bar chart into vertical bar chart by using `coord_flip()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()  
```

## Changing the y- and x-axis range

The scatterplot on the right is slightly misleading because the y-aixs and x-axis range are not equal.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)   
```

The code chunk below fixed both the y-axis and x-axis range from 0-100.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))    
```

# Themes

Themes control elements of the graph not related to the data. For example:

-   background colour

-   size of fonts

-   gridlines

-   colour of labels

Built-in themes include: - `theme_gray()` (default) - `theme_bw()` - `theme_classic()`

A list of theme can be found at this [link](https://ggplot2.tidyverse.org/reference/ggtheme.html). Each theme element can be conceived of as either a line (e.g. x-axis), a rectangle (e.g. graph background), or text (e.g. axis title).

## Working with theme

The code chunk below plot a horizontal bar chart using `theme_gray()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()     
```

A horizontal bar chart plotted using `theme_classic()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()      
```

A horizontal bar chart plotted using `theme_minimal()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()       
```
