---
title: "Take-home Exercise 1: Visual Analytics of Factors Influencing Student Academic Performance at a Vietnamese University"
author: "Walaiporn Pawitranon"
format: 
  html:
    code-fold: false
execute: 
  warning: false
  error: true
  eval: true
---

# 1. Overview

## 1.1 Background

This dataset is based on a survey of 2,170 students and alumni from the University of Education, Vietnam National University, Hanoi, conducted between March and May 2023. It includes information on students’ GPA, demographic characteristics, learning motivation, university support, lecturer support, facilities, training programs, and peer relationships.

Academic performance is influenced by both personal factors (such as motivation and learning adaptation) and external factors (such as teaching quality, institutional support, and family background). In the context of educational innovation in Vietnam, where learners are placed at the center of the learning process, understanding these influences is important for improving student outcomes.

Although the data is self-reported, it provides valuable insights into students’ learning experiences and their relationship with GPA, making it suitable for exploratory data analysis and visual analytics in educational research.

## 1.2 The data

The dataset consists of 2,170 responses from students and alumni of the University of Education, Vietnam National University, Hanoi, with 22 variables capturing demographic, social, and academic factors. The data was collected and reported in [this study](https://www.sciencedirect.com/science/article/pii/S2352340925001702). Below are some key highlights:

Numerical Features include GPA and time-related variables such as average hours spent studying, using social media, and socializing with friends. These variables allow quantitative assessment of student behaviors and learning outcomes.

Categorical Features cover student demographics (year of study, gender, minority status, policy-supported students, economic background), parents’ education and occupation, and perceptions of university and lecturer support, curriculum suitability, class competition, and peer influence. These variables provide insights into factors that may influence academic performance and motivation.

Overall, the dataset offers a comprehensive view of student characteristics, behaviors, and support systems, making it suitable for exploring relationships between various internal and external factors and learning outcomes.

## 1.3 The task

Using the 2023 Student Questionnaire Dataset from the University of Education, Vietnam National University, Hanoi, this post aims to use appropriate Exploratory Data Analysis (EDA) methods and ggplot2 functions to uncover:

-   The distribution of students’ learning outcomes, measured by GPA, and

-   The relationship between GPA and students’ demographic, personal, and external factors, including year of study, gender, family background, time spent studying, adaptation to university, learning methods, university support, lecturer support, classroom competition, and peer influence.

The goal is to identify key factors that influence student performance and provide meaningful insights for educational research and intervention planning.

# 2. Getting start

## 2.1 Loading packages

+---------------+---------------------------------------------------------------------------------------------------------------+
| Library       | Description                                                                                                   |
+===============+===============================================================================================================+
| **tidyverse** | Core packages for data wrangling, filtering, summarizing, and plotting (`dplyr`, `ggplot2`, `readr`, `tidyr`) |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **readxl**    | Read Excel `.xlsx` files                                                                                      |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **ggplot2**   |                                                                                                               |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **scales**    | Format axis labels, percentages, and numbers for readability                                                  |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **plotly**    | Create interactive scatterplots and coordinated multiple views                                                |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **ggthemes**  | A ggplot extension that provides the user additional themes, scales, and geoms for the ggplots package        |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **patchwork** | Combine multiple ggplot2 plots into one figure                                                                |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **ggpubr**    | Publication-ready plots and statistical annotation                                                            |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **ggtext**    | Enhanced text rendering for ggplot2 labels and tooltips                                                       |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **ggdist**    | Visualisations of distributions and uncertainty in ggplot2                                                    |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **ggridges**  |                                                                                                               |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **skimr**     | Summarize dataset with descriptive statistics for EDA                                                         |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **janitor**   | Clean column names, tabulate categorical variables for EDA                                                    |
+---------------+---------------------------------------------------------------------------------------------------------------+

The code chunk below will be used load these R packages into RStudio environment

```{r}
pacman::p_load(tidyverse, readxl, ggiraph, ggthemes, ggdist, plotly, crosstalk, 
  DT, patchwork, scales, ggpubr, ggtext, skimr, janitor)
```

## 2.2 Data import

Database paper.xlsx will be used.

In the code chunk below, read_xlsx() of readr package is used to import Database paper.xlsx into R and saved it into a tibble data.frame.

```{r}
student_data <- read_xlsx("data/Database paper.xlsx")
```

# 3. Data wrangling

## 3.1 Variables selection

For this post, only the columns relevant to our planned visualizations were retained. The dataset’s accompanying Codebook was particularly helpful in understanding the meaning of each variable, their coding system, and the types of responses collected.

The retained variables can be grouped into four broad categories, capturing aspects of University Support, Gender/Personal Demographics, Socio-economic Status, and Student Behavior/Performance. Details are extracted from the Codebook:

+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Category              | Variable name      | Question                                                                     | Encoding scheme                     |
+=======================+====================+==============================================================================+=====================================+
| Personal Demographics | Year               | What year student are you in?                                                | 1 - First-year                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Second-year                     |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Third-year                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Fourth-year                     |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Graduated                       |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Personal Demographics | Gender             | What is your gender?                                                         | 1 - Male                            |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Female                          |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Socio-economic Status | Policy_Stu         | Are you a student on policy government support?                              | 1 - Yes                             |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - No                              |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Socio-economic Status | Poor_Stu           | Is your family a poor household?                                             | 1 - Yes                             |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - No                              |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Family Background     | Father_Edu         | What educational background of your father?                                  | 1 - Primary school                  |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Secondary school                |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - High school                     |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - College school                  |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - University/graduate education   |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 6 - Other                           |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Family Background     | Mother_Edu         | What educational background of your mother?                                  | 1 - Primary school                  |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Secondary school                |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - High school                     |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - College school                  |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - University/graduate education   |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 6 - Other                           |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Student Behavior      | Time_Friends       | How much time do you spend hanging out with your friends on average per day? | 1 - under 1 hours                   |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - from 1 to lower 2 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - from 2 to lower 3 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - from 3 to lower 4 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - over 4 hour                     |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Student Behavior      | Time_SocicalMedia  | How much time do you spend using social media on average per day?            | 1 - under 1 hours                   |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - from 1 to lower 2 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - from 2 to lower 3 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - from 3 to lower 4 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - over 4 hour                     |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Student Behavior      | Time_Studying      | How much time do you spend studying on average per day?                      | 1 - under 2 hours                   |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - from 2 to lower 4 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - from 4 to lower 6 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - from 6 to lower 8 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - over 8 hour                     |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Performance           | GPA                | What is your cumulative grade point average (GPA)?                           | 1 - Under 2.0 – Poor                |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - From 2.0 to lower 2.5 – Average |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - From 2.5 to lower 3.2 – Fair    |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - From 3.2 to lower  3.6 – Good   |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Over 3.6 – Excellent            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| University Support    | Adapt_Learning_Uni | How level are you adapting to the learning environment at the university?    | 1 - Not at all                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Little                          |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Moderate                        |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Quite                           |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Very                            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| University Support    | Study_Methods      | How level are your studying methods affecting your outcomes?                 | 1 - Not at all                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Little                          |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Moderate                        |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Quite                           |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Very                            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| University Support    | SupportOf_Uni      | How level is the university's support for you?                               | 1 - Not at all                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Little                          |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Moderate                        |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Quite                           |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Very                            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| University Support    | SupportOf_Lec      | How level is the lecturer's support (affecting your outcomes)?               | 1 - Not at all                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Little                          |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Moderate                        |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Quite                           |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Very                            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| University Support    | Facilitie_Uni      | How responsive are the university facilities?                                | 1 - Not at all                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Little                          |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Moderate                        |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Quite                           |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Very                            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| University Support    | TrainingCurriculum | How level is suitable of the university curriculum?                          | 1 - Not at all                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Little                          |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Moderate                        |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Quite                           |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Very                            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+

glimpse() of the dplyr package allows us to see all columns and their data type in the data frame.

```{r}
glimpse(student_data)
```


A new dataset, student_viz_data, was generated by selecting only the variables relevant to the planned visualizations.

```{r}

selected_vars <- c(
  "Year", "Gender", "Poor_Stu", "Policy_Stu", "Father_Edu", "Mother_Edu",
  "Time_Friends", "Time_SocicalMedia", "Time_Studying",
  "GPA", "Adapt_Learning_Uni", "Study_Methods", "Quality_Lecturer",
  "SupportOf_Uni", "SupportOf_Lec", "Facilitie_Uni", "TrainingCurriculum","Competitive_Class", "InfuenceF_Friends"
)

student_viz_data <- student_data %>%
  select(all_of(selected_vars))

head(student_viz_data)

```

## 3.2 Handling missing value

To check the total number of missing values in the whole dataset, the code chunk below was used.

```{r}
sum(is.na(student_viz_data))
```

To examine missing values for each variable individually, the code chunk below was used to compute the number of missing observations by column.

```{r}
colSums(is.na(student_viz_data))
```

After inspecting the dataset using colSums(is.na(student_data)), no missing values were detected across all variables. This indicates that the dataset is complete and does not require further treatment such as row deletion, variable removal, or imputation. Therefore, all observations were retained for subsequent data preparation, analysis, and visualization.

## 3.3 Recoding categorical variables

Although many variables in the dataset are recorded using numerical codes (e.g., 1–5), these values do not represent actual quantities or measurements but instead correspond to ordered categories or levels of perception. If left as numeric variables, R would treat them as continuous numbers, implying equal intervals between categories and allowing inappropriate mathematical operations such as averaging or regression as if they were true numerical measurements. This can lead to misleading results and incorrect visual interpretations. Therefore, these variables need to be recoded as ordered factors with meaningful labels to ensure accurate and appropriate analysis and visualization by using the code chunk below.

```{r}
library(dplyr)

student_viz_data <- student_viz_data |>
  mutate(
    # ----- Student profile -----
    Year = factor(Year,
                  levels = c(1,2,3,4,5),
                  labels = c("First-year",
                             "Second-year",
                             "Third-year",
                             "Fourth-year",
                             "Graduated")),

    Gender = factor(Gender,
                    levels = c(1,2),
                    labels = c("Male", "Female")),

    Poor_Stu = factor(Poor_Stu,
                      levels = c(1,2),
                      labels = c("Poor household", "Not poor")),

    Policy_Stu = factor(Policy_Stu,
                        levels = c(1,2),
                        labels = c("Policy supported", "Not supported")),

    # ----- Parents' education -----
    Father_Edu = factor(Father_Edu,
                        levels = 1:6,
                        labels = c("Primary school",
                                   "Secondary school",
                                   "High school",
                                   "College",
                                   "University/Graduate",
                                   "Other")),

    Mother_Edu = factor(Mother_Edu,
                        levels = 1:6,
                        labels = c("Primary school",
                                   "Secondary school",
                                   "High school",
                                   "College",
                                   "University/Graduate",
                                   "Other")),

    # ----- Time use -----
    Time_Friends = factor(Time_Friends,
                          levels = c(1,2,3,4,5),
                          labels = c("Under 1 hour",
                                     "1–<2 hours",
                                     "2–<3 hours",
                                     "3–<4 hours",
                                     "Over 4 hours")),

    Time_SocicalMedia = factor(Time_SocicalMedia,
                               levels = c(1,2,3,4,5),
                               labels = c("Under 1 hour",
                                          "1–<2 hours",
                                          "2–<3 hours",
                                          "3–<4 hours",
                                          "Over 4 hours")),

    Time_Studying = factor(Time_Studying,
                           levels = c(1,2,3,4,5),
                           labels = c("Under 2 hours",
                                      "2–<4 hours",
                                      "4–<6 hours",
                                      "6–<8 hours",
                                      "Over 8 hours")),

    # ----- Academic outcome -----
    GPA = factor(GPA,
                 levels = c(1,2,3,4,5),
                 labels = c("Under 2.0 – Poor",
                            "2.0–<2.5 – Average",
                            "2.5–<3.2 – Fair",
                            "3.2–<3.6 – Good",
                            "Over 3.6 – Excellent")),

    # ----- University support & learning -----
    Adapt_Learning_Uni = factor(Adapt_Learning_Uni,
                                levels = 1:5,
                                labels = c("Not at all",
                                           "Little",
                                           "Moderate",
                                           "Quite",
                                           "Very")),

    Study_Methods = factor(Study_Methods,
                           levels = 1:5,
                           labels = c("Not at all",
                                      "Little",
                                      "Moderate",
                                      "Quite",
                                      "Very")),

    SupportOf_Uni = factor(SupportOf_Uni,
                           levels = 1:5,
                           labels = c("Not at all",
                                      "Little",
                                      "Moderate",
                                      "Quite",
                                      "Very")),

    SupportOf_Lec = factor(SupportOf_Lec,
                           levels = 1:5,
                           labels = c("Not at all",
                                      "Little",
                                      "Moderate",
                                      "Quite",
                                      "Very")),

    Facilitie_Uni = factor(Facilitie_Uni,
                           levels = 1:5,
                           labels = c("Not at all",
                                      "Little",
                                      "Moderate",
                                      "Quite",
                                      "Very")),
    
    Quality_Lecturer = factor(Quality_Lecturer,
                           levels = 1:5,
                           labels = c("Not at all",
                                      "Little",
                                      "Moderate",
                                      "Quite",
                                      "Very")),

    TrainingCurriculum = factor(TrainingCurriculum,
                                levels = 1:5,
                                labels = c("Not at all",
                                           "Little",
                                           "Moderate",
                                           "Quite",
                                           "Very")),
    Competitive_Class = factor(Competitive_Class,
                                levels = 1:5,
                                labels = c("Not at all",
                                           "Little",
                                           "Moderate",
                                           "Quite",
                                           "Very")),
    InfuenceF_Friends = factor(InfuenceF_Friends,
                                levels = 1:5,
                                labels = c("Not at all",
                                           "Little",
                                           "Moderate",
                                           "Quite",
                                           "Very"))
  )

```

After converting the selected variables to factors, we used glimpse() to quickly inspect the structure of the dataset and confirm that all categorical variables were correctly recoded.

```{r}
glimpse(student_viz_data)
```

# 4. Exploratory Data Analysis

## Insight 1: Student Demographic

::: panel-tabset
## Number of Students by Year

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"

ggplot(student_viz_data, aes(x = factor(Year), fill = factor(Year))) +
  geom_bar() +
  labs(title = "Number of Students by Year",
       x = "Year of Study",
       y = "Count of Students") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

## Gender Distribution of Students

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"
ggplot(student_viz_data, aes(x = Gender, fill = Gender)) +
  geom_bar() +
  labs(title = "Gender Distribution of Students",
       x = "Gender",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("skyblue", "pink"))
```

## Students from Poor Households

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"
ggplot(student_viz_data, aes(x = Poor_Stu, fill = Poor_Stu)) +
  geom_bar() +
  labs(title = "Students from Poor Households",
       x = "Poor Student (1=Yes, 2=No)",
       y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Paired")

```

## Students under Government Policy Support

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"
ggplot(student_viz_data, aes(x = Policy_Stu, fill = Policy_Stu)) +
  geom_bar() +
  labs(title = "Students under Government Policy Support",
       x = "Policy Supported (1=Yes, 2=No)",
       y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")

```
:::

::: {.callout-tip title="Insight"}
Overall, the visualizations suggest that:

:::

## Insight 2: Student Time Allocation

To examine the relationship between students’ time allocation and academic performance (GPA), three separate visualizations were created using jittered scatter plots in `ggplot2`. The three plots explored:

-    Time spent studying vs. GPA

-    Time spent with friends vs. GPA

-    Time spent on social media vs. GPA

In each plot, GPA was treated as the dependent variable (y-axis), while the respective time variable (categorical hours groups) was used as the independent variable (x-axis).

A jittered scatter plot (`geom_jitter`) was applied to reduce overplotting and allow individual data points to be visible, especially where many students shared the same GPA category and time group. This helps reveal the distribution of students within each time category rather than just showing aggregated values..

::: panel-tabset
## Relationship between Daily Study Time and GPA

-   Higher study time categories tend to have higher average GPA (as shown by the black diamond).

<!-- -->

-   You might also see that very few students study over 8 hours, but those who do generally achieve high GPA.

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"
ggplot(student_viz_data, aes(x = Time_Studying, y = GPA, color = Time_Studying)) +
  geom_jitter(width = 0.2, height = 0.2, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4, fill = "black") +
  labs(
    title = "Relationship between Daily Study Time and GPA",
    x = "Time Studying (hours category)",
    y = "GPA"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Relationship between Time Spent with Friends and GPA

-   Moderate time with friends might not negatively affect GPA.

<!-- -->

-   Extremely high socializing hours (4+ hours) could correlate with slightly lower GPA.

<!-- -->

-   Some students maintain high GPA even with high social time, showing individual differences.

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"
ggplot(student_viz_data, aes(x = Time_Friends, y = GPA, color = Time_Friends)) +
  geom_jitter(width = 0.2, height = 0.2, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4, fill = "black") +
  labs(
    title = "Relationship between Time Spent with Friends and GPA",
    x = "Time with Friends (hours category)",
    y = "GPA"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Relationship between Social Media Usage and GPA

-   Excessive social media use (3–4+ hours) may correlate with lower GPA averages.

<!-- -->

-   Low-to-moderate social media use seems compatible with higher GPA.

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"
ggplot(student_viz_data, aes(x = Time_SocicalMedia, y = GPA, color = Time_SocicalMedia)) +
  geom_jitter(width = 0.2, height = 0.2, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4, fill = "black") +
  labs(
    title = "Relationship between Social Media Usage and GPA",
    x = "Time on Social Media (hours category)",
    y = "GPA"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
:::

::: {.callout-tip title="Insight"}
Overall, the three visualizations suggest that:

-   Students’ behaviors (study, socializing, social media) show **different impacts on GPA**.

<!-- -->

-   Study time is most positively correlated with GPA.

<!-- -->

-   Time with friends shows little clear relationship with GPA.

<!-- -->

-   Social media time shows a mild negative relationship with GPA.

This indicates that how students allocate their time matters, with academic-focused activities (studying) being more strongly associated with better performance than leisure activities.
:::

## Insight 3: Student Study Time vs Performance

To examine whether study time is associated with GPA, two complementary visual approaches were used: (1) a descriptive visualization and (2) an inferential visualization.

First, a descriptive bar chart was created using ggplot2 from the tidyverse to show the distribution of students across five study-time categories. The code used geom_bar() with after_stat(prop) to display percentages rather than raw counts, allowing comparison of how common each study-time category was among respondents. The variable Time_Studying was treated as a factor to reflect its Likert-scale (ordinal) nature.

Second, an inferential visualization was produced using the ggscatterstats() function from the ggstatsplot package. Because both Study Time and GPA are measured on Likert (ordinal) scales, a non-parametric Spearman rank correlation (type = "np") was applied rather than a Pearson correlation. Marginal histograms were removed (marginal = FALSE) since the variables are discrete categories rather than continuous measures. This plot visually displays the relationship between study time and GPA while embedding the Spearman correlation coefficient and p-value directly in the graph.

```{r}
p1 <- ggplot(student_viz_data, aes(x = factor(Time_Studying))) +
  geom_bar(aes(y = after_stat(prop), group = 1)) +
  scale_y_continuous(labels = scales::percent) +
  labs(
  title = "Distribution of Daily Study Time",
    x = "Time Spent Studying",
    y = "Percentage of Students"
  ) +
  theme_minimal(base_size = 10)+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 11)
  ) 

p2 <- ggscatterstats(
  data = student_viz_data,
  x = Time_Studying,
  y = GPA,
  type = "np",
  marginal = FALSE,
  title = "Association between Study Time and GPA",
  xlab = "Time Spent Studying",
  ylab = "GPA Level"
) +
  theme_minimal(base_size = 10) +   # cleaner text
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 11)
  )

(p1 | p2) + 
  plot_layout(widths = c(1.5, 2.3)) +
  plot_annotation(
    title = "Study Time and GPA Analysis"
  )


```

::: {.callout-tip title="Insight"}
The descriptive bar chart shows that the majority of students reported studying in the highest category (“over 8 hours”), indicating that most respondents are relatively intensive study. Very few students reported studying less than 2 hours per day.

However, the inferential scatter plot with Spearman correlation suggests that increased study time does not show a strong or meaningful monotonic association with GPA in this sample (ρ ≈ 0, p ≈ 1.00). This indicates that although many students study for long hours, longer study time alone does not necessarily translate into higher GPA.

Together, these two plots suggest that academic performance is likely influenced by factors beyond study duration alone, such as study methods, lecturer quality, university support, or peer influence.
:::


## Insight 4: Relationship of Class Competitiveness by Friends' Influence

A diverging stacked bar chart was used to compare the distribution of classroom competitiveness (Competitive_Class) between students who reported positive friends’ influence and those who did not. Percentages were calculated within each group to allow fair comparison.


::: panel-tabset
## A plot

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"
library(dplyr)
library(tidyr)

div_data <- student_viz_data |>
  count(InfuenceF_Friends, Competitive_Class) |>
  group_by(InfuenceF_Friends) |>
  mutate(perc = n / sum(n) * 100) |>
  ungroup()

div_data1 <- student_viz_data |>
  count(InfuenceF_Friends, GPA) |>
  group_by(InfuenceF_Friends) |>
  mutate(perc = n / sum(n) * 100) |>
  ungroup()

div_data2 <- student_viz_data |>
  count(Competitive_Class, GPA) |>
  group_by(Competitive_Class) |>
  mutate(perc = n / sum(n) * 100) |>
  ungroup()
```

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"
library(dplyr)
library(tidyr)

ggplot(div_data, 
       aes(x = perc,
           y = InfuenceF_Friends,
           fill = Competitive_Class)) +
  geom_col(position = position_stack(vjust = 0.5)) +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Class Competitiveness by Friends' Influence",
    x = "Percentage of Students",
    y = "Friends have positive influence?"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

```

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"
ggplot(div_data1, 
       aes(x = perc,
           y = InfuenceF_Friends,
           fill = GPA)) +
  geom_col(position = position_stack(vjust = 0.5)) +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Class Competitiveness by Friends' Influence",
    x = "Percentage of Students",
    y = "Friends have positive influence?"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

```
```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"
ggplot(div_data2, 
       aes(x = perc,
           y = Competitive_Class,
           fill = GPA)) +
  geom_col(position = position_stack(vjust = 0.5)) +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Class Competitiveness by Friends' Influence",
    x = "Percentage of Students",
    y = "Friends have positive influence?"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

```



## b

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"

```

## c

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code"

```
:::

::: {.callout-tip title="Insight"}
The visualization shows that students who perceive positive influence from friends are more likely to report being in highly competitive classes compared to those who do not. Conversely, students who do not feel positively influenced by friends tend to have a higher proportion of low or moderate class competitiveness. This suggests that peer influence and perceived academic environment are related and may jointly shape students’ learning experience.

:::

## Insight 5: Family background Equity & opportunity

```{r}
library(ggplot2)
library(dplyr)
library(ggthemes)

# 1️⃣ Father's Education vs GPA
ggplot(student_viz_data, aes(x = factor(Father_Edu), y = GPA, fill = factor(Father_Edu))) +
  geom_boxplot() +
  stat_summary(fun=mean, geom="point", shape=23, size=3, color="black", fill="yellow") +
  labs(title="GPA by Father's Education Level",
       x="Father's Education",
       y="GPA") +
  theme_minimal() +
  scale_fill_brewer(palette="Pastel2")

# 2️⃣ Mother's Education vs GPA
ggplot(student_viz_data, aes(x = factor(Mother_Edu), y = GPA, fill = factor(Mother_Edu))) +
  geom_boxplot() +
  stat_summary(fun=mean, geom="point", shape=23, size=3, color="black", fill="yellow") +
  labs(title="GPA by Mother's Education Level",
       x="Mother's Education",
       y="GPA") +
  theme_minimal() +
  scale_fill_brewer(palette="Pastel2")

# 3️⃣ Economic Background (Poor vs Not Poor) vs GPA
ggplot(student_viz_data, aes(x = factor(Poor_Stu), y = GPA, fill = factor(Poor_Stu))) +
  geom_boxplot() +
  stat_summary(fun=mean, geom="point", shape=23, size=3, color="black", fill="yellow") +
  labs(title="GPA by Economic Background",
       x="Poor Student (1=Yes, 2=No)",
       y="GPA") +
  theme_minimal() +
  scale_fill_brewer(palette="Paired")

```
```{r}
ggplot(student_viz_data, aes(x = GPA, y = factor(Father_Edu), fill = factor(Father_Edu))) + geom_density_ridges(alpha = 0.7, scale = 1.2, rel_min_height = 0.01) + geom_jitter(height = 0.1, width = 0.1, alpha = 0.5) + labs(title = "GPA Distribution by Father's Education", x = "GPA", y = "Father's Education") + theme_minimal() + scale_fill_brewer(palette = "Set3")
```

```{r}
likert_data <- student_data %>%
  select(
    Adapt_Learning_Uni, Study_Methods, SupportOf_Uni,
    SupportOf_Lec, Facilitie_Uni, Quality_Lecturer,
    TrainingCurriculum, Competitive_Class, InfuenceF_Friends
  ) %>%
  pivot_longer(cols = everything(),
               names_to = "Question",
               values_to = "Response")

ggplot(likert_data, aes(x = factor(Response))) +
  geom_bar(aes(y = after_stat(prop), group = 1)) +
  facet_wrap(~ Question) +
  scale_y_continuous(labels = scales::percent)

```

::: {.callout-tip title="Insight"}
Overall, the visualizations suggest that:

:::

## Insight 6: University support

To examine the relationship between students’ perceived university support (SupportOf_Uni) and lecturer support (SupportOf_Lec), a stacked bar chart with embedded inferential statistics was produced using the ggbarstats() function from the ggstatsplot package.

The stacked bar chart provides a descriptive visualization of how responses to lecturer support are distributed within each level of university support. Percentages within each bar were displayed to allow easy comparison across categories rather than relying on raw counts alone.

To complement the descriptive visualization, an inferential test of association was automatically performed by ggbarstats(). Because both variables are categorical Likert-scale responses, a Pearson Chi-square test of independence was used to assess whether the two variables are statistically associated. In addition, Cramér’s V was reported as a measure of effect size to indicate the strength of the relationship.

::: panel-tabset
## The plot

```{r}
#| echo: false
library(ggstatsplot)
ggbarstats(
  data = student_viz_data,
  x = SupportOf_Uni,
  y = SupportOf_Lec
)

```

## The code chunk

```{r}
#| results: false
library(ggstatsplot)
ggbarstats(
  data = student_viz_data,
  x = SupportOf_Uni,
  y = SupportOf_Lec
)

```
:::

::: {.callout-tip title="Insight"}

-   **Pearson χ² = 1795.83, p \< 0.001** → the relationship is statistically significant.

-   **Cramér’s V ≈ 0.45** → this represents a **moderate to strong association** between university support and lecturer support.

The stacked bar chart and Pearson Chi-square test show a strong and significant association between perceived university support and lecturer support (Cramér’s V ≈ 0.45). Students who feel highly supported by the university are also much more likely to perceive strong support from their lecturers, indicating that institutional and instructional support are closely interconnected in shaping student experience.
:::



# 5. Summary

# 6. Reference

<https://www.sciencedirect.com/science/article/pii/S2352340925001702>

<https://quarto.org/docs/computations/execution-options.html>
