---
title: "Take-home Exercise 1: Visual Analytics of Factors Influencing Student Academic Performance at a Vietnamese University"
author: "Walaiporn Pawitranon"
execute: 
  warning: false
  echo: true
  eval: true
---

# 1. Overview

## 1.1 Background

This dataset is based on a survey of 2,170 students and alumni from the University of Education, Vietnam National University, Hanoi, conducted between March and May 2023. It includes information on students’ GPA, demographic characteristics, learning motivation, university support, lecturer support, facilities, training programs, and peer relationships.

Academic performance is influenced by both personal factors (such as motivation and learning adaptation) and external factors (such as teaching quality, institutional support, and family background). In the context of educational innovation in Vietnam, where learners are placed at the center of the learning process, understanding these influences is important for improving student outcomes.

Although the data is self-reported, it provides valuable insights into students’ learning experiences and their relationship with GPA, making it suitable for exploratory data analysis and visual analytics in educational research.

## 1.2 The data

The dataset consists of 2,170 responses from students and alumni of the University of Education, Vietnam National University, Hanoi, with 22 variables capturing demographic, social, and academic factors. The data was collected and reported in [this study](https://www.sciencedirect.com/science/article/pii/S2352340925001702). Below are some key highlights:

Numerical Features include GPA and time-related variables such as average hours spent studying, using social media, and socializing with friends. These variables allow quantitative assessment of student behaviors and learning outcomes.

Categorical Features cover student demographics (year of study, gender, minority status, policy-supported students, economic background), parents’ education and occupation, and perceptions of university and lecturer support, curriculum suitability, class competition, and peer influence. These variables provide insights into factors that may influence academic performance and motivation.

Overall, the dataset offers a comprehensive view of student characteristics, behaviors, and support systems, making it suitable for exploring relationships between various internal and external factors and learning outcomes.

## 1.3 The task

Using the 2023 Student Questionnaire Dataset from the University of Education, Vietnam National University, Hanoi, this post aims to use appropriate Exploratory Data Analysis (EDA) methods and ggplot2 functions to uncover:

-   The distribution of students’ learning outcomes, measured by GPA, and

-   The relationship between GPA and students’ demographic, personal, and external factors, including year of study, gender, family background, time spent studying, adaptation to university, learning methods, university support, lecturer support, classroom competition, and peer influence.

The goal is to identify key factors that influence student performance and provide meaningful insights for educational research and intervention planning.

# 2. Getting start

## 2.1 Loading packages

+---------------+---------------------------------------------------------------------------------------------------------------+
| Library       | Description                                                                                                   |
+===============+===============================================================================================================+
| **tidyverse** | Core packages for data wrangling, filtering, summarizing, and plotting (`dplyr`, `ggplot2`, `readr`, `tidyr`) |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **readxl**    | Read Excel `.xlsx` files                                                                                      |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **ggplot2**   |                                                                                                               |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **scales**    | Format axis labels, percentages, and numbers for readability                                                  |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **plotly**    | Create interactive scatterplots and coordinated multiple views                                                |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **ggthemes**  | A ggplot extension that provides the user additional themes, scales, and geoms for the ggplots package        |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **patchwork** | Combine multiple ggplot2 plots into one figure                                                                |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **ggpubr**    | Publication-ready plots and statistical annotation                                                            |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **ggtext**    | Enhanced text rendering for ggplot2 labels and tooltips                                                       |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **ggdist**    | Visualisations of distributions and uncertainty in ggplot2                                                    |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **ggridges**  |                                                                                                               |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **skimr**     | Summarize dataset with descriptive statistics for EDA                                                         |
+---------------+---------------------------------------------------------------------------------------------------------------+
| **janitor**   | Clean column names, tabulate categorical variables for EDA                                                    |
+---------------+---------------------------------------------------------------------------------------------------------------+

The code chunk below will be used load these R packages into RStudio environment

```{r}
pacman::p_load(tidyverse, readxl, ggiraph, ggthemes, ggdist, plotly, crosstalk, 
  DT, patchwork, scales, ggpubr, ggtext, skimr, janitor)
```

## 2.2 Data import

Database paper.xlsx will be used.

In the code chunk below, read_xlsx() of readr package is used to import Database paper.xlsx into R and saved it into a tibble data.frame.

```{r}
student_data <- read_xlsx("data/Database paper.xlsx")
```

# 3. Data wrangling

## 3.1 Variables selection

For this post, only the columns relevant to our planned visualizations were retained. The dataset’s accompanying Codebook was particularly helpful in understanding the meaning of each variable, their coding system, and the types of responses collected.

The retained variables can be grouped into four broad categories, capturing aspects of University Support, Gender/Personal Demographics, Socio-economic Status, and Student Behavior/Performance. Details are extracted from the Codebook:

+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Category              | Variable name      | Question                                                                     | Encoding scheme                     |
+=======================+====================+==============================================================================+=====================================+
| Personal Demographics | Year               | What year student are you in?                                                | 1 - First-year                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Second-year                     |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Third-year                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Fourth-year                     |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Graduated                       |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Personal Demographics | Gender             | What is your gender?                                                         | 1 - Male                            |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Female                          |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Socio-economic Status | Policy_Stu         | Are you a student on policy government support?                              | 1 - Yes                             |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - No                              |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Socio-economic Status | Poor_Stu           | Is your family a poor household?                                             | 1 - Yes                             |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - No                              |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Family Background     | Father_Edu         | What educational background of your father?                                  | 1 - Primary school                  |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Secondary school                |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - High school                     |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - College school                  |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - University/graduate education   |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 6 - Other                           |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Family Background     | Mother_Edu         | What educational background of your mother?                                  | 1 - Primary school                  |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Secondary school                |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - High school                     |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - College school                  |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - University/graduate education   |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 6 - Other                           |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Student Behavior      | Time_Friends       | How much time do you spend hanging out with your friends on average per day? | 1 - under 1 hours                   |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - from 1 to lower 2 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - from 2 to lower 3 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - from 3 to lower 4 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - over 4 hour                     |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Student Behavior      | Time_SocicalMedia  | How much time do you spend using social media on average per day?            | 1 - under 1 hours                   |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - from 1 to lower 2 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - from 2 to lower 3 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - from 3 to lower 4 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - over 4 hour                     |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Student Behavior      | Time_Studying      | How much time do you spend studying on average per day?                      | 1 - under 2 hours                   |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - from 2 to lower 4 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - from 4 to lower 6 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - from 6 to lower 8 hours         |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - over 8 hour                     |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| Performance           | GPA                | What is your cumulative grade point average (GPA)?                           | 1 - Under 2.0 – Poor                |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - From 2.0 to lower 2.5 – Average |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - From 2.5 to lower 3.2 – Fair    |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - From 3.2 to lower  3.6 – Good   |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Over 3.6 – Excellent            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| University Support    | Adapt_Learning_Uni | How level are you adapting to the learning environment at the university?    | 1 - Not at all                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Little                          |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Moderate                        |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Quite                           |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Very                            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| University Support    | Study_Methods      | How level are your studying methods affecting your outcomes?                 | 1 - Not at all                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Little                          |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Moderate                        |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Quite                           |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Very                            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| University Support    | SupportOf_Uni      | How level is the university's support for you?                               | 1 - Not at all                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Little                          |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Moderate                        |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Quite                           |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Very                            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| University Support    | SupportOf_Lec      | How level is the lecturer's support (affecting your outcomes)?               | 1 - Not at all                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Little                          |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Moderate                        |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Quite                           |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Very                            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| University Support    | Facilitie_Uni      | How responsive are the university facilities?                                | 1 - Not at all                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Little                          |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Moderate                        |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Quite                           |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Very                            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+
| University Support    | TrainingCurriculum | How level is suitable of the university curriculum?                          | 1 - Not at all                      |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 2 - Little                          |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 3 - Moderate                        |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 4 - Quite                           |
|                       |                    |                                                                              |                                     |
|                       |                    |                                                                              | 5 - Very                            |
+-----------------------+--------------------+------------------------------------------------------------------------------+-------------------------------------+

glimpse() of the dplyr package allows us to see all columns and their data type in the data frame.

```{r}
glimpse(student_data)
```

```{r}
summary(student_data)
```

A new dataset, student_viz_data, was generated by selecting only the variables relevant to the planned visualizations.

```{r}

selected_vars <- c(
  "Year", "Gender", "Poor_Stu", "Policy_Stu", "Father_Edu", "Mother_Edu",
  "Time_Friends", "Time_SocicalMedia", "Time_Studying",
  "GPA", "Adapt_Learning_Uni", "Study_Methods",
  "SupportOf_Uni", "SupportOf_Lec", "Facilitie_Uni", "TrainingCurriculum"
)

student_viz_data <- student_data %>%
  select(all_of(selected_vars))

head(student_viz_data)

```

## 3.2 Handling missing value

To check the total number of missing values in the whole dataset, the code chunk below was used.

```{r}
sum(is.na(student_viz_data))
```

To examine missing values for each variable individually, the code chunk below was used to compute the number of missing observations by column.

```{r}
colSums(is.na(student_viz_data))
```

After inspecting the dataset using colSums(is.na(student_data)), no missing values were detected across all variables. This indicates that the dataset is complete and does not require further treatment such as row deletion, variable removal, or imputation. Therefore, all observations were retained for subsequent data preparation, analysis, and visualization.

## 3.3 Recoding categorical variables

Although many variables in the dataset are recorded using numerical codes (e.g., 1–5), these values do not represent actual quantities or measurements but instead correspond to ordered categories or levels of perception. If left as numeric variables, R would treat them as continuous numbers, implying equal intervals between categories and allowing inappropriate mathematical operations such as averaging or regression as if they were true numerical measurements. This can lead to misleading results and incorrect visual interpretations. Therefore, these variables need to be recoded as ordered factors with meaningful labels to ensure accurate and appropriate analysis and visualization by using the code chunk below.

```{r}
library(dplyr)

student_viz_data <- student_viz_data |>
  mutate(
    # ----- Student profile -----
    Year = factor(Year,
                  levels = c(1,2,3,4,5),
                  labels = c("First-year",
                             "Second-year",
                             "Third-year",
                             "Fourth-year",
                             "Graduated")),

    Gender = factor(Gender,
                    levels = c(1,2),
                    labels = c("Male", "Female")),

    Poor_Stu = factor(Poor_Stu,
                      levels = c(1,2),
                      labels = c("Poor household", "Not poor")),

    Policy_Stu = factor(Policy_Stu,
                        levels = c(1,2),
                        labels = c("Policy supported", "Not supported")),

    # ----- Parents' education -----
    Father_Edu = factor(Father_Edu,
                        levels = 1:6,
                        labels = c("Primary school",
                                   "Secondary school",
                                   "High school",
                                   "College",
                                   "University/Graduate",
                                   "Other")),

    Mother_Edu = factor(Mother_Edu,
                        levels = 1:6,
                        labels = c("Primary school",
                                   "Secondary school",
                                   "High school",
                                   "College",
                                   "University/Graduate",
                                   "Other")),

    # ----- Time use -----
    Time_Friends = factor(Time_Friends,
                          levels = c(1,2,3,4,5),
                          labels = c("Under 1 hour",
                                     "1–<2 hours",
                                     "2–<3 hours",
                                     "3–<4 hours",
                                     "Over 4 hours")),

    Time_SocicalMedia = factor(Time_SocicalMedia,
                               levels = c(1,2,3,4,5),
                               labels = c("Under 1 hour",
                                          "1–<2 hours",
                                          "2–<3 hours",
                                          "3–<4 hours",
                                          "Over 4 hours")),

    Time_Studying = factor(Time_Studying,
                           levels = c(1,2,3,4,5),
                           labels = c("Under 2 hours",
                                      "2–<4 hours",
                                      "4–<6 hours",
                                      "6–<8 hours",
                                      "Over 8 hours")),

    # ----- Academic outcome -----
    GPA = factor(GPA,
                 levels = c(1,2,3,4,5),
                 labels = c("Under 2.0 – Poor",
                            "2.0–<2.5 – Average",
                            "2.5–<3.2 – Fair",
                            "3.2–<3.6 – Good",
                            "Over 3.6 – Excellent")),

    # ----- University support & learning -----
    Adapt_Learning_Uni = factor(Adapt_Learning_Uni,
                                levels = 1:5,
                                labels = c("Not at all",
                                           "Little",
                                           "Moderate",
                                           "Quite",
                                           "Very")),

    Study_Methods = factor(Study_Methods,
                           levels = 1:5,
                           labels = c("Not at all",
                                      "Little",
                                      "Moderate",
                                      "Quite",
                                      "Very")),

    SupportOf_Uni = factor(SupportOf_Uni,
                           levels = 1:5,
                           labels = c("Not at all",
                                      "Little",
                                      "Moderate",
                                      "Quite",
                                      "Very")),

    SupportOf_Lec = factor(SupportOf_Lec,
                           levels = 1:5,
                           labels = c("Not at all",
                                      "Little",
                                      "Moderate",
                                      "Quite",
                                      "Very")),

    Facilitie_Uni = factor(Facilitie_Uni,
                           levels = 1:5,
                           labels = c("Not at all",
                                      "Little",
                                      "Moderate",
                                      "Quite",
                                      "Very")),

    TrainingCurriculum = factor(TrainingCurriculum,
                                levels = 1:5,
                                labels = c("Not at all",
                                           "Little",
                                           "Moderate",
                                           "Quite",
                                           "Very"))
  )

```

After converting the selected variables to factors, we used glimpse() to quickly inspect the structure of the dataset and confirm that all categorical variables were correctly recoded.

```{r}
glimpse(student_viz_data)
```

# 4. Univariate Analysis


# 5. Exploratory Data Analysis

EDA1: Student demograpic

```{r}
library(ggplot2)
library(dplyr)
library(ggthemes)

# Count of students by Year
ggplot(student_viz_data, aes(x = factor(Year), fill = factor(Year))) +
  geom_bar() +
  labs(title = "Number of Students by Year",
       x = "Year of Study",
       y = "Count of Students") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")

# Gender composition
ggplot(student_viz_data, aes(x = Gender, fill = Gender)) +
  geom_bar() +
  labs(title = "Gender Distribution of Students",
       x = "Gender",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("skyblue", "pink"))

# Economic status (Poor vs Not Poor)
ggplot(student_viz_data, aes(x = Poor_Stu, fill = Poor_Stu)) +
  geom_bar() +
  labs(title = "Students from Poor Households",
       x = "Poor Student (1=Yes, 2=No)",
       y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Paired")

# Policy-supported students
ggplot(student_viz_data, aes(x = Policy_Stu, fill = Policy_Stu)) +
  geom_bar() +
  labs(title = "Students under Government Policy Support",
       x = "Policy Supported (1=Yes, 2=No)",
       y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")

```

# 6. Summary

# 

# 7. Reference

## 
